### 2025-05-14 chapters/mpif/ms.Rnw
Finished dissertation

A common approach to modeling nonlinear dynamic systems is through the use of mechanistic models \citep{auger21}.
These models involve the proposal of a system of equations that describe how the dynamic system state evolves stochastically over time.
When combined with a model relating the state to observable quantities, we obtain a partially observed Markov process (POMP) model, also known as a state space model or a hidden Markov model.
Once calibrated to data, such models provide a quantitative description of the observations while simultaneously adhering to a given scientific hypothesis about how the data are generated.
Despite the growing ability of deep learning and other advanced machine learning techniques to extract useful information from complex data sets, mechanistic models continue to be important in modern science since mechanistic models are superior to machine learning models for interpretability of parameters \citep{baker18,hogg24}.
The ability of mechanistic models to describe scientifically important yet unobservable system states \citep{andrade22} also facilitates predictions of the impact of interventions on the system, such as masking or vaccination efforts during a disease outbreak \citep{wheeler24}.

@@

To set up mathematical notation for PanelPOMP models, we first give a formal definition of a POMP.
A POMP model comprises an unobserved Markov process $\mathbf{X}(t)$, where $t \in \mathcal{T}\subset \mathbb{R}$ and $\mathbf{X}(t)$ takes values in $\mathcal{X} \subset \mathbb{R}^{d_x}$.
The process is initialized at time $t_0 \in \mathcal{T}$, such that $t_0\leq t'$ for all $t' \in \mathcal{T}$.
Observations are drawn from the dynamic system at time points $t_1, \ldots, t_N \in \mathcal{T}$, and we denote the observation measured at time $t_n$ as $\mathbf{y}^*_n \in \mathcal{Y} \subset \mathbb{R}^{d_y}$.
We assume that the observation $\mathbf{y}^*_n$ is a single realization of a random variable $\mathbf{Y}_n$.

While $\mathbf{X}(t)$ may be a continuous or discrete time process, the value of $\mathbf{X}(t)$ at observation times $t_{1:N}$ is often of particular interest, and so we write $\mathbf{X}_n = \mathbf{X}(t_n)$.
This gives rise to a joint probability density or probability mass function $f_{\mathbf{X}_{0:N}, \mathbf{Y}_{1:N}}(\mathbf{x}_{0:N}, \mathbf{y}_{1:N}; \theta)$ at the observation times $t_{0:N}$.
For the remainder of this article, we use the term {\it density} to refer to both probability densities and probability mass functions, the latter being formally treated as densities with respect to a counting measure on a finite or countable set.
To complete the POMP specification, we assume that the observed random variables $\mathbf{Y}_{1:N}$ are conditionally independent given the values of the latent process $\mathbf{X}_{1:N}$.

@@

Our goal is to make inferences about the parameter $\theta$ by maximizing the likelihood function
\begin{align}
L(\theta;\mathbf{y}^*) &= \int f_{\mathbf{X}_{0:N}, \mathbf{Y}_{1:N}}(\mathbf{x}_{0:N}, \mathbf{y}^*_{1:N}; \theta)\, d\mathbf{x}_{0:N}. \label{eq:likelihood}
\end{align}
Although the unit densities in Eq.~\ref{eq:ppomp} can be factored as a result of the dynamic independence between measurement units, a defining feature of a panel model is that the parameter vector $\theta$ remains relevant across all the dynamic systems.
Therefore, inference methodology used to obtain estimates of $\theta$ should use data from each unit in the PanelPOMP.
A special case of particular interest arises when $\theta = (\phi, \psi_{1:U})$, where $\phi \in \Theta_\phi \subset \mathbb{R}^{d_\phi}$ is a vector of parameters that are shared across each unit, and $\psi_{u} \in \Theta_\psi \subset \mathbb{R}^{d_u}$ are parameters that are specific to unit $u$, formally,
$$
f_{X_{u, 0:N_u}, Y_{u,1:N_u}}(x_{u, 0:N_u}, y_{u, 1:N}; \theta) = f_{X_{u, 0:N_u}, Y_{u,1:N_u}}(x_{u, 0:N_u}, y_{u, 1:N}; \phi, \psi_u).
$$
Algorithm~\ref{alg:mpif} describes an iterated filtering algorithms for PanelPOMP models that obtains the MLE for both shared $(\phi)$ and unit specific $(\psi_u)$ parameters.
With $\mathrm{MARGINALIZE}=\mathrm{FALSE}$, this is the PIF algorithm of \citet{breto20}.
Our innovation occurs when $\mathrm{MARGINALIZE}=\mathrm{TRUE}$, and we call this the  marginalized panel iterated filter (MPIF).
MPIF can be decomposed into sub-steps that each resemble a single iteration of the IF2 algorithm.
However, MPIF adds an additional step to IF2 by marginalizing out the unit-specific parameters that are irrelevant for the unit currently being filtered.
Since the parameter distributions are represented via Monte Carlo samples, the marginalization step is implemented by not resampling the particles corresponding to the unit-specific parameters of other units.
In other words, instead of resampling all parameter particles in line~\ref{mpif:update} of Algorithm~\ref{alg:mpif}, we only resample particles for the shared parameters and the unit-specific parameters related to the data of the specific unit under consideration.

@@

\begin{theorem}\label{theorem:pif}
  Consider a PanelPOMP model defined by Eq.~\ref{eq:ppomp}.
  Assume conditions \ref{assumption:1}--\ref{assumption:kernel} of Appendix~\ref{sec:assumptions} are satisfied, and write $\hat{\theta} = \argmax_{\theta} L(\theta;\mathbf{y}^*)$ from Eq.~\ref{eq:likelihood}.
  Denote the output of Algorithm~\ref{alg:mpif} without marginalization as $\Theta_{1:J}^{(M)}$.
  Then there exists some positive sequences $\{C_M\}_{M \geq 1}$ and $\{\epsilon_M\}_{M \geq 1}$ where $\lim_{M \rightarrow \infty}\epsilon_M = 0$ such that for all $(J, M) \in \mathbb{N}^2$,
  $$
  E\bigg[\Big|\frac{1}{J}\sum_{i=1}^J \Theta_j^{(M)} - \hat{\theta}\Big|_2\bigg] \leq \frac{C_M}{\sqrt{J}} + \epsilon_M
  $$
\end{theorem}
\begin{proof}[Proof outline]
  This theorem is an extension of Theorem~4 of \citet{chen24} to PanelPOMP models, following the approach of \citet{breto20}, who showed that PanelPOMP models can be expressed as a lower-dimensional POMP model. 
  Using this technique, the lower-dimensional version of the model has a corresponding self-organized state space model that has a filtering distribution with mass centered at $\hat{\theta}$, and Algorithm~\ref{alg:mpif} is a particle filter applied to the self-organized state space model. A full proof is provided in Appendix~\ref{sec:panelTheory}.
\end{proof}

@@

If we denote $\pi_i(\theta)$ as the posterior distribution of the parameter vector $\theta$ after the $i$th Bayesian update, and $\mathbf{y}^* = \mathbf{y}_{1:N}^*$ as the observed data, we can express an iterated Bayesian update as the following:
\begin{align*}
\pi_1(\theta) &\propto f(\mathbf{y}^* ;\theta)\pi_0(\theta), \\
\pi_2(\theta) &\propto f(\mathbf{y}^* ;\theta)\pi_1(\theta) \propto f^2(\mathbf{y}^* ;\theta)\pi_0(\theta),\\
&\vdots \\
\pi_m(\theta) &\propto f^m(\mathbf{y}^* ;\theta)\pi_0(\theta).
\end{align*}
In this representation, $f(\mathbf{y}^*; \theta)$ is the likelihood function (Eq.~\ref{eq:likelihood}), and $\pi_0(\theta)$ is the original prior distribution for $\theta$.
If we let $m\rightarrow \infty$, the effect of the initial prior distribution diminishes, and the $m$th posterior has all of its mass centered at the MLE.
This can be shown by taking the limit of $\pi_m(\theta) / \pi_m(\hat{\theta})$ as $m$ goes to infinity: if $\theta = \hat{\theta}$, then the limit is one, and zero otherwise \citep{lele07}.

@@

\begin{corollary}\label{corollary:perturbed}
  Consider the setup of Theorem~\ref{theorem:GG}. If the parameter perturbations are Gaussian with covariance matrix $\sigma^2_m \Sigma_0$ for some initial covariance matrix $\Sigma_0 \in \R^{(U+1) \times (U+1)}$ and sequence $\sigma^2_m = o(1/m)$, then the $m$th iteration of Eqs.~\ref{eq:margBayesPerturb} and \ref{eq:MPIFupdatePerturb} corresponds to a Gaussian distribution with mean $\mu'_m \in \R^{U+1}$ and covariance $\Sigma'_m \in \R^{U+1\times U+1}$. If $\theta^* = (\phi^*, \psi^*_1, \ldots, \psi^*_{U})$ denotes the MLE, then $\|\mu'_m - \theta^*\|_2 \rightarrow 0$ and $\|\Sigma'_m\|_2 \rightarrow 0$.
\end{corollary}

@@

ggplot(df_all_mu) +
  geom_point(
    aes(x = mu1, y = mu2, col = as.factor(step)),
    pch = 18, size = 1
  ) +
  geom_path(
    data = df_all_Sigma,
    aes(x = X1, y = X2, col = as.factor(step))
  ) +
  geom_path(
    data = like_elipse,
    aes(x = X1, y = X2), linetype = 'dashed'
  ) +
  geom_point(x = 0, y = 0, col = 'red', pch = 'x') +
  theme_bw() +
  scale_x_continuous(limits = c(-3.12, 13.5)) +
  scale_y_continuous(limits = c(-3.12, 13.5)) +
  theme(aspect.ratio = 1) +
  labs(color = 'Step') +
  scale_color_manual(
    values = c(
      "#800026",
      "#bd0026",
      "#e31a1c",
      "#fc4e2a",
      "#fd8d3c",
      "#feb24c"
    )
  ) +
  labs(x = TeX('$\\phi$'), y = TeX('$\\psi$')) +
  theme(legend.position = 'right') +
  facet_wrap(~group, labeller = labeller(group = method_names)) + 
  theme(plot.margin = unit(c(0, 0, 0, 0), 'cm'))
@
\caption{\label{fig:DC}Data Cloning and Marginalized Data Cloning for two parameter model with Gaussian likelihoods and priors. The ellipses show the region of the parameter distribution that contains $95\%$ of the probability mass of the distribution. The black dashed line shows this region for the likelihood surface, and the red ``x" marks the MLE.}
\end{figure}

@@

cowplot::plot_grid(depletion_1, depletion_2, nrow = 2, labels = "AUTO", rel_heights = c(0.425, 0.575))
@
\caption{\label{fig:depletion}Updating parameter distributions with a single $u = 1$ iteration of both versions of Algorithm~\ref{alg:mpif}. (A) The total number of unique particles representing each parameter. The dashed horizontal line shows that MPIF maintains the number of particles for $\Psi_2$ over time. (B) Parameter particle swarm of a single update with and without marginalization compared to the true posterior distribution.}
\end{figure}

@@

\noindent where $K_u$ is the carrying capacity for the population in unit $u$, $r_u$ is a positive parameter that corresponds to the growth rate, and $\epsilon_{u, n}$ are iid. lognormal random variables such that $\log \epsilon_{u, n} \sim \normal[0, \sigma_u^2]$.
Measurements of the population are obtained via the density

@@

cowplot::plot_grid(gg_smallU, gg_bigU, nrow = 2, rel_heights = c(0.4, 0.6))
@
\caption{\label{fig:gomp}Comparison of the MPIF and PIF algorithms for fitting the stochastic Gompertz population model. 
The solid horizontal line shows the true maximum likelihood, determined via the Kalman filter and a numeric optimizer, an intractable approach for high-dimensional parameter spaces $(U > 100)$, so in these cases the dashed line indicates the likelihood at the data-generating parameters. 
Each algorithm used 50 unique starting points.
Vertical bars span the tenth percentile to the maximum likelihood values; lowest values were excluded due to poor likelihoods from the PIF algorithm.}
\end{figure}

\section{Example: Measles in the United Kingdom}

@@

print(gg)
@
\caption{\label{fig:measles1} Log-likelihoods yielded by the PIF and MPIF algorithms for the mechanistic measles model to the UK data. 
Rows correspond to a different number of particles $J$ used in Algorithm~\ref{alg:mpif}. 
The log-likelihood is evaluated at iterations 20, 56, 92, 128, 164, and 200, marked by vertical grey lines.}
\end{figure}

>>>>

A common approach to modeling nonlinear dynamic systems is through the use of mechanistic models \citep{auger21}.
These models involve the proposal of a system of equations that describe how the dynamic system state evolves stochastically over time.
When combined with a model relating the state to observable quantities, we obtain a POMP model, described in Chapter~\ref{chpt:intro}.
Once calibrated to data, such models provide a quantitative description of the observations while simultaneously adhering to a given scientific hypothesis about how the data are generated.
Despite the growing ability of deep learning and other advanced machine learning techniques to extract useful information from complex data sets, mechanistic models continue to be important in modern science since mechanistic models are superior to machine learning models for interpretability of parameters \citep{baker18,hogg24}.
The ability of mechanistic models to describe scientifically important yet unobservable system states \citep{andrade22} also facilitates predictions of the impact of interventions on the system, such as masking or vaccination efforts during a disease outbreak \citep{wheeler24}.

@@

% To set up mathematical notation for PanelPOMP models, we first give a formal definition of a POMP.
% A POMP model comprises an unobserved Markov process $\mathbf{X}(t)$, where $t \in \mathcal{T}\subset \mathbb{R}$ and $\mathbf{X}(t)$ takes values in $\mathcal{X} \subset \mathbb{R}^{d_x}$.
% The process is initialized at time $t_0 \in \mathcal{T}$, such that $t_0\leq t'$ for all $t' \in \mathcal{T}$.
% Observations are drawn from the dynamic system at time points $t_1, \ldots, t_N \in \mathcal{T}$, and we denote the observation measured at time $t_n$ as $\mathbf{y}^*_n \in \mathcal{Y} \subset \mathbb{R}^{d_y}$.
% We assume that the observation $\mathbf{y}^*_n$ is a single realization of a random variable $\mathbf{Y}_n$.
% 
% While $\mathbf{X}(t)$ may be a continuous or discrete time process, the value of $\mathbf{X}(t)$ at observation times $t_{1:N}$ is often of particular interest, and so we write $\mathbf{X}_n = \mathbf{X}(t_n)$.
% This gives rise to a joint probability density or probability mass function $f_{\mathbf{X}_{0:N}, \mathbf{Y}_{1:N}}(\mathbf{x}_{0:N}, \mathbf{y}_{1:N}; \theta)$ at the observation times $t_{0:N}$.
% For the remainder of this article, we use the term {\it density} to refer to both probability densities and probability mass functions, the latter being formally treated as densities with respect to a counting measure on a finite or countable set.
% To complete the POMP specification, we assume that the observed random variables $\mathbf{Y}_{1:N}$ are conditionally independent given the values of the latent process $\mathbf{X}_{1:N}$.

@@

Our goal is to make inferences about the parameter $\theta$ by maximizing the likelihood function (Eq.~\ref{eq:likedef}).
% \begin{align}
% \math(\theta;\mathbf{y}^*) &= \int f_{\mathbf{X}_{0:N}, \mathbf{Y}_{1:N}}(\mathbf{x}_{0:N}, \mathbf{y}^*_{1:N}; \theta)\, d\mathbf{x}_{0:N}. \label{eq:likelihood}
% \end{align}
Although the unit densities in Eq.~\ref{eq:ppomp} can be factored as a result of the dynamic independence between measurement units, a defining feature of a panel model is that the parameter vector $\theta$ remains relevant across all the dynamic systems.
Therefore, inference methodology used to obtain estimates of $\theta$ should use data from each unit in the PanelPOMP.
A special case of particular interest arises when $\theta = (\phi, \psi_{1:U})$, where $\phi \in \Theta_\phi \subset \mathbb{R}^{d_\phi}$ is a vector of parameters that are shared across each unit, and $\psi_{u} \in \Theta_\psi \subset \mathbb{R}^{d_u}$ are parameters that are specific to unit $u$, formally,
$$
f_{X_{u, 0:N_u}, Y_{u,1:N_u}}(x_{u, 0:N_u}, y_{u, 1:N}; \theta) = f_{X_{u, 0:N_u}, Y_{u,1:N_u}}(x_{u, 0:N_u}, y_{u, 1:N}; \phi, \psi_u).
$$
Algorithm~\ref{alg:mpif} describes an iterated filtering algorithms for PanelPOMP models that obtains the MLE for both shared $(\phi)$ and unit specific $(\psi_u)$ parameters.
With $\mathrm{MARGINALIZE}=\mathrm{FALSE}$, this is the PIF algorithm of \citet{breto20}.
Our innovation occurs when $\mathrm{MARGINALIZE}=\mathrm{TRUE}$, and we call this the  marginalized panel iterated filter (MPIF).
MPIF can be decomposed into sub-steps that each resemble a single iteration of the IF2 algorithm.
However, MPIF adds an additional step to IF2 by marginalizing out the unit-specific parameters that are irrelevant for the unit currently being filtered.
Since the parameter distributions are represented via Monte Carlo samples, the marginalization step is implemented by not resampling the particles corresponding to the unit-specific parameters of other units.
In other words, instead of resampling all parameter particles in line~\ref{mpif:update} of Algorithm~\ref{alg:mpif}, we only resample particles for the shared parameters and the unit-specific parameters related to the data of the specific unit under consideration.

@@

\begin{theorem}\label{theorem:pif}
  Consider a PanelPOMP model defined by Eq.~\ref{eq:ppomp}.
  Assume conditions \ref{assumption:1}--\ref{assumption:kernel} of Appendix~\ref{sec:assumptions} are satisfied, and write $\hat{\theta} = \argmax_{\theta} L(\theta;\mathbf{y}^*)$ from Eq.~\ref{eq:likedef}.
  Denote the output of Algorithm~\ref{alg:mpif} without marginalization as $\Theta_{1:J}^{(M)}$.
  Then there exists some positive sequences $\{C_M\}_{M \geq 1}$ and $\{\epsilon_M\}_{M \geq 1}$ where $\lim_{M \rightarrow \infty}\epsilon_M = 0$ such that for all $(J, M) \in \mathbb{N}^2$,
  $$
  E\bigg[\Big\vert\frac{1}{J}\sum_{i=1}^J \Theta_j^{(M)} - \hat{\theta}\Big\vert_2\bigg] \leq \frac{C_M}{\sqrt{J}} + \epsilon_M
  $$
\end{theorem}
\begin{proof}[Proof outline]
  This theorem is an extension of Theorem~4 of \citet{chen24} to PanelPOMP models, following the approach of \citet{breto20}, who showed that PanelPOMP models can be expressed as a lower-dimensional POMP model. 
  Using this technique, the lower-dimensional version of the model has a corresponding self-organized state space model that has a filtering distribution with mass centered at $\hat{\theta}$, and Algorithm~\ref{alg:mpif} is a particle filter applied to the self-organized state space model. A full proof is provided in Appendix~\ref{sec:panelTheory}.
\end{proof}

@@

If we denote $\pi_i(\theta)$ as the posterior distribution of the parameter vector $\theta$ after the $i$th Bayesian update, and $\mathbf{y}^* = \mathbf{y}_{1:N}^*$ as the observed data, we can express an iterated Bayesian update as the following:
\begin{align*}
\pi_1(\theta) &\propto f(\mathbf{y}^* ;\theta)\pi_0(\theta), \\
\pi_2(\theta) &\propto f(\mathbf{y}^* ;\theta)\pi_1(\theta) \propto f^2(\mathbf{y}^* ;\theta)\pi_0(\theta),\\
&\vdots \\
\pi_m(\theta) &\propto f^m(\mathbf{y}^* ;\theta)\pi_0(\theta).
\end{align*}
In this representation, $f(\mathbf{y}^*; \theta)$ is the likelihood function (Eq.~\ref{eq:likedef}), and $\pi_0(\theta)$ is the original prior distribution for $\theta$.
If we let $m\rightarrow \infty$, the effect of the initial prior distribution diminishes, and the $m$th posterior has all of its mass centered at the MLE.
This can be shown by taking the limit of $\pi_m(\theta) / \pi_m(\hat{\theta})$ as $m$ goes to infinity: if $\theta = \hat{\theta}$, then the limit is one, and zero otherwise \citep{lele07}.

@@

\begin{corollary}\label{corollary:perturbed}
  Consider the setup of Theorem~\ref{theorem:GG}. If the parameter perturbations are Gaussian with covariance matrix $\sigma^2_m \Sigma_0$ for some initial covariance matrix $\Sigma_0 \in \R^{(U+1) \times (U+1)}$ and sequence $\sigma^2_m = o(1/m)$, then the $m$th iteration of Eqs.~\ref{eq:margBayesPerturb} and \ref{eq:MPIFupdatePerturb} corresponds to a Gaussian distribution with mean $\mu'_m \in \R^{U+1}$ and covariance $\Sigma'_m \in \R^{U+1\times U+1}$. If $\theta^* = (\phi^*, \psi^*_1, \ldots, \psi^*_{U})$ denotes the MLE, then $\vert\mu'_m - \theta^*\vert_2 \rightarrow 0$ and $\|\Sigma'_m\|_2 \rightarrow 0$.
\end{corollary}

@@

ggplot(df_all_mu) +
  geom_point(
    aes(x = mu1, y = mu2, col = as.factor(step)),
    pch = 18, size = 1
  ) +
  geom_path(
    data = df_all_Sigma,
    aes(x = X1, y = X2, col = as.factor(step))
  ) +
  geom_path(
    data = like_elipse,
    aes(x = X1, y = X2), linetype = 'dashed'
  ) +
  geom_point(x = 0, y = 0, col = 'red', pch = 'x') +
  theme_bw() +
  scale_x_continuous(limits = c(-3.12, 13.5)) +
  scale_y_continuous(limits = c(-3.12, 13.5)) +
  theme(aspect.ratio = 1) +
  labs(color = 'Step') +
  scale_color_manual(
    values = c(
      "#800026",
      "#bd0026",
      "#e31a1c",
      "#fc4e2a",
      "#fd8d3c",
      "#feb24c"
    )
  ) +
  labs(x = TeX('$\\phi$'), y = TeX('$\\psi$')) +
  theme(legend.position = 'right') +
  facet_wrap(~group, labeller = labeller(group = method_names)) + 
  theme(plot.margin = unit(c(0, 0, 0, 0), 'cm'))
@
\caption[Data cloning compared to marginalized data cloning for Gaussian densities.]{\label{fig:DC}Data cloning and Marginalized data cloning for two parameter model with Gaussian likelihoods and priors. The ellipses show the region of the parameter distribution that contains $95\%$ of the probability mass of the distribution. The black dashed line shows this region for the likelihood surface, and the red ``x" marks the MLE.}
\end{figure}

@@

cowplot::plot_grid(depletion_1, depletion_2, nrow = 2, labels = "AUTO", rel_heights = c(0.425, 0.575))
@
\caption[Single unit iteration of MPIF on toy model.]{\label{fig:depletion}Updating parameter distributions with a single $u = 1$ iteration of both versions of Algorithm~\ref{alg:mpif}. (A) The total number of unique particles representing each parameter. The dashed horizontal line shows that MPIF maintains the number of particles for $\Psi_2$ over time. (B) Parameter particle swarm of a single update with and without marginalization compared to the true posterior distribution.}
\end{figure}

@@

\noindent where $K_u$ is the carrying capacity for the population in unit $u$, $r_u$ is a positive parameter that corresponds to the growth rate, and $\epsilon_{u, n}$ are iid. log-normal random variables such that $\log \epsilon_{u, n} \sim \normal[0, \sigma_u^2]$.
Measurements of the population are obtained via the density

@@

cowplot::plot_grid(gg_smallU, gg_bigU, nrow = 2, rel_heights = c(0.4, 0.6))
@
\caption[MPIF compared to PIF on Gompertz population model.]{\label{fig:gomp}Comparison of the MPIF and PIF algorithms for fitting the stochastic Gompertz population model. 
The solid horizontal line shows the true maximum likelihood, determined via the Kalman filter and a numeric optimizer, an intractable approach for high-dimensional parameter spaces $(U > 100)$, so in these cases the dashed line indicates the likelihood at the data-generating parameters. 
Each algorithm used 50 unique starting points.
Vertical bars span the tenth percentile to the maximum likelihood values; lowest values were excluded due to poor likelihoods from the PIF algorithm.}
\end{figure}

\section{Example: Measles in the United Kingdom}\label{sec:UKmeas}

@@

print(gg)
@
\caption[MPIF comparison using mechanistic model for UK measles incidence data.]{\label{fig:measles1} Log-likelihoods yielded by the PIF and MPIF algorithms for the mechanistic measles model to the UK data. 
Rows correspond to a different number of particles $J$ used in Algorithm~\ref{alg:mpif}. 
The log-likelihood is evaluated at iterations 20, 56, 92, 128, 164, and 200, marked by vertical grey lines.}
\end{figure}

### 2025-05-14 front_materials/acknowledgments.tex
Finished dissertation

I am deeply grateful to all of the mentors in my life who have helped me get to this point.
I am particularly grateful for my advisor, Dr. Edward L. Ionides, who has provided invaluable support for me throughout my academic journey and has helped shape my beliefs about statistics and higher education.
I wish to also thank the other members of my committee who have provided their advise and mentorship. 
Dr. Aaron A. King

Most importantly I would like to thank my spouse, Haylee Wheeler, who has been incredibly supportive and encouranging though my entire academic career.
The progress that I have made as a scholar would not have been possible without her.
You have always been willing to listen to my complaints and struggles, and been with me to celebrate life's victories.

>>>>

I am deeply grateful to all of the friends, family, and mentors in my life who have helped me get to this point.
I will name some of the key contributors here, recognizing that many others have also played an important role in my journey, even if not mentioned individually.

First, I would like to express my heartfelt gratitude to my adviser, Dr. Edward L. Ionides.
I could not have asked for a better mentor, and I am immensely thankful for all you have done for me.
Beyond guiding me through my academic journey, you have profoundly influenced my beliefs and perspectives on statistics and higher education.
You have always been readily available for discussions about projects or concerns, and I have consistently looked forward to our meetings.
As I advance in my own career, I will often reflect on the exemplary standard you set as both a researcher and mentor.

I would also like to thank Dr. Aaron A. King for his mentorship. 
Software development became a much more significant part of my experience at the University of Michigan than I initially anticipated, and you played a crucial role in helping me develop good software practices. 
Whether through one-on-one interactions or by providing numerous examples of well-designed, documented, and reproducible code, your guidance has been instrumental.
You have always been available when I needed advice and were quick to offer helpful insights.

I wish to also thank the other members of my committee, Dr. Kerby Shedden and Dr. Jeffrey Regier, who have provided their advise and mentorship.
Though we did not meet regularly, I am incredibly grateful for the time you have spent helping me prepare this thesis, provide letters of recommendation, and provide general career advice.
I have looked to both of you as exemplary scholars and mentors, and you have helped shape my experience as a student more than you may realize.

I've also been fortunate to have a wonderful group of friends and classmates who have supported me throughout my PhD studies.
I would be remiss not to specifically thank members of my cohort: Prayag Chatha, Pramit Das, Luke Francisco, Kihyuk Hong, Declan McNamara, Bo Meng, Gang Qiao, Junting Wang, and Yilei Zhang.
We began the PhD program amid a global pandemic, which presented a number of unique challenges.
Together, you helped me navigate the virtual classroom environment, collaborated on homework and group projects, and provided invaluable support over the years.
This group of individuals contains many of the smartest and most talented people I have ever met, and I hope our paths will once again cross in the near future.

My family deserves a great deal of thanks for their unwavering support over the years.
To my father, you have been the best role model, showing me what it means to be a good person.
As long as I can remember, I aspired to be just like you, this perhaps being my earliest motivation to pursue a PhD.
To my mother, thank you for your endless love and support.
You have been a role model for me in countless ways. 
I am continually amazed by how you manage to balance a meaningful career, social life, personal hobbies, and a healthy lifestyle, all while prioritizing your family above everything else.
Your energy, kindness, and talents are truly unmatched.

Most importantly I would like to thank my spouse, Haylee Wheeler, who has been incredibly supportive and encouraging throughout my entire academic career.
The progress that I have made as a scholar would not have been possible without her.
Despite not having any connections to Michigan, you were willing to move across the country with me just so that I could pursue my dream of getting a PhD.
When we arrived in Ann Arbor in the midst of the pandemic, we had little opportunity to meet people or make friends; I'm am not confident that I could have endured this stage of my PhD program without you.
You have always been willing to listen to my complaints and struggles, and been with me to celebrate life's victories.
I'm so grateful for your love and support in ways that I cannot possibly express.
You are the best friend and companion I could have ever hoped for, and I am excited to share the next chapter of our lives together.

This dissertation contains three projects described in Chapters~\ref{chpt:arima}--\ref{chpt:mpif}.
The work outlined in each of the chapters have either been published in a peer-reviewed journal (Chapter~\ref{chpt:haiti}), or are currently being prepared for publication (Chapters~\ref{chpt:arima} and \ref{chpt:mpif}). 
These projects include work that was done in conjunction with the coauthors of these papers, and I would like to explicitly thank them and highlight their contributions.
\begin{itemize}
  \item  In Chapter~\ref{chpt:arima}, my adviser Edward L. Ionides met with me weekly to discuss ideas and progress related to this project.
  He also provided valuable feedback and suggestions on how to present my algorithm and results, and was instrumental in refining the manuscript text. 
  Importantly, he was the first to notice the apparent optimization issue present in existing algorithms, which was the original motivation for this work.
  \item Chapter~\ref{chpt:haiti} was a large undertaking by several members of Dr. Ionides's research group, and I had the privilege of supervising a group of undergraduate students (AnnaElaine Rosengart, Kevin Tan, and Noah Treutle) as well as a master student (Zhuoxun (Josh) Jiang) who worked on this project. 
  All coauthors were part of weekly discussions on the project and helped refine the manuscript text.
  AnnaElaine had a major role in writing software to build Model~1, and Josh wrote code that enabled us to reproduce existing results that were obtained using Model~2.
  Kevin Tan and Noah Treutle began investigations related to the linear reductions in transmission rate for Model~1.
  This work is now published in PLoS Computational Biology \citep{wheeler24}. 
  \item Chapter~\ref{chpt:mpif} represents joint work with Dr. Ionides and Aaron Abkemeier, a classmate and coauthor of this project. 
  Aaron is an expert in Measles transmission models, and is primarily responsible for the analysis presented in Section~\ref{sec:UKmeas}.
  I am incredibly grateful for his contributions, and the chance I had to work closely with him on this project.
\end{itemize}

I would also like to acknowledge funding opportunities that have helped make this work possible.
First, I am incredibly honored to have received a Rackham Merit Fellowship (RMF) that has supported me throughout my PhD experience.
I also received funds through my Advisor from a National Science Foundation Research Training Grant (RTG) DMS-1646108, and also received a Rackham Research Grant (U086898) that helped fund the projects reported in this thesis.
